---
title: "Final Project"
author: "Alicia Helfrich, Olivia Gomez, and Akbar Naqvi"
format: html
editor: visual
embed-resources: true
---

Outline from meeting:

-   Lit review

-   Overview of data sources

-   Exploratory Data Analysis and Data Cleaning

-   Building Models

-   Analyzing results - variable importance

-   Conclusion Limitations

-   download file for data folder:

(Data and Documents - Primary Data File)  links: - https://www.census.gov/programs-surveys/sipp/data/datasets/2022-data/2022.html - https://www.census.gov/programs-surveys/sipp/data/datasets/2019-data/2019.html

## Project Overview

-   OG: state categories of eligibility rules 
-   OG: intro draft

## Data Sources

-   OG: look into survey methodology 

```{r}

library(tidyverse)
library(haven)
library(data.table)

```



### Reading in Data 
```{r}

data_2019 <- fread("data/pu2019.csv", 
                   sep = "|",
                   showProgress = TRUE)

data_2019_jan <- data_2019 %>%
  filter(MONTHCODE == 1) %>%
  # filter out people without Medicaid in the month of January (using EMDMTH)
  
  # this might be the outcome: RMARKTPLACE
  
# model will help target outreach by predicting those most likely to churn
  
# adding variable of types of eligibility rules by state (grouping)
  
# two tables 


```

## Exploratory Data Analysis
### missingness (Akbar)
### correlation (Alicia)
### explore categorical variables of interest (Alicia - depending on Aaron's response)
### explore the distribution of outcome variable 

### question for Aaron - does reducing categorical variables matter as much for tree models?
### how do we decide what variables to drop?  

## Predictive Modeling

## Analyzing Results

## Conclusions and Limitations
